---
title: "Introduction to R"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

In its most basic form, R can be used as a simple calculator. Consider the following arithmetic operators:

* Addition: `+`
* Subtraction: `-`
* Multiplication: `*`
* Division: `/`
* Exponentiation: `^`
* Modulo: `%%`

The last two might need some explaining:

* The `^` operator raises the number to its left to the power of the number to its right: for example `3^2` is 9.
* The modulo returns the remainder of the division of the number to the left by the number on its right, for example 5 modulo 3 or `5 %% 3` is 2.

```{r Arithmetic with R}
# An addition
5 + 5 
#> [1] 10

# A subtraction
5 - 5 
#> [1] 0

# A multiplication
3 * 5
#> [1] 15

 # A division
(5 + 5) / 2 
#> [1] 5

# Exponentiation
2^5
#> [1] 32

# Modulo
28%%6
#> [1] 4

```
<sup>Created on 2022-01-17 by the [reprex package](https://reprex.tidyverse.org) (v2.0.1)</sup>

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
