---
title: "Intermediate Importing Data in R"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
```

# 1. Importing data from databases (Part 1)
## Establish a connection

The first step to import data from a SQL database is creating a connection to it. As Filip explained, you need different packages depending on the database you want to connect to. All of these packages do this in a uniform way, as specified in the `DBI` package.

<a href="http://www.rdocumentation.org/packages/DBI/functions/dbConnect" target="_blank" rel="noopener noreferrer">`dbConnect()`</a> creates a connection between your R session and a SQL database. The first argument has to be a `DBIdriver` object, that specifies how connections are made and how data is mapped between R and the database. Specifically for MySQL databases, you can build such a driver with <a href="http://www.rdocumentation.org/packages/RMySQL/functions/MySQLDriver-class" target="_blank" rel="noopener noreferrer">`RMySQL::MySQL()`</a>. 

If the MySQL database is a remote database hosted on a server, you'll also have to specify the following arguments in <a href="http://www.rdocumentation.org/packages/DBI/functions/dbConnect" target="_blank" rel="noopener noreferrer">`dbConnect()`</a>: `dbname`, `host`, `port`, `user` and `password`. Most of these details have already been provided.


**Instructions:**

* Load the `DBI` library, which is already installed on DataCamp's servers.
* Edit the <a href="http://www.rdocumentation.org/packages/DBI/functions/dbConnect" target="_blank" rel="noopener noreferrer">`dbConnect()`</a> call to connect to the MySQL database. Change the `port` argument (`3306`) and `user` argument (`"student"`).

**Solution:**

```{r,eval=TRUE}
# Load the DBI package
library(DBI)

# Edit dbConnect() call
con <- dbConnect(RMySQL::MySQL(), 
                 dbname = "tweater", 
                 host = "courses.csrrinzqubik.us-east-1.rds.amazonaws.com", 
                 port = 3306,
                 user = "student",
                 password = "datacamp")
```



## Inspect the connection

Now that you've successfully created the database connection, let's have a closer look at it. With the object `con` available in your workspace, can you tell which of the following statements is true?

**Possible Answers:**

:white_large_square: con is an SQLConnection object.<br>
:white_large_square: con is a PostgreSQLConnection object.<br>
:white_check_mark: con is an MySQLConnection object.<br>
:white_large_square: con is an NoSQLConnection object.<br>


## List the database tables

After you've successfully connected to a remote MySQL database, the next step is to see what tables the database contains. You can do this with the <a href="http://www.rdocumentation.org/packages/DBI/functions/dbListTables" target="_blank" rel="noopener noreferrer">`dbListTables()`</a> function. As you might remember from the video, this function requires the connection object as an input, and outputs a character vector with the table names.


**Instructions:**

* Add code to create a vector `tables`, that contains the tables in the tweater database. You can connect to this database through the `con` object.
* Display the structure of `tables`; what's the class of this vector?

**Solution:**

```{r,eval=TRUE}
# Build a vector of table names: tables
tables <- dbListTables(con)

# Display structure of tables
str(tables)
```



## Import users

As you might have guessed by now, the database contains data on a more tasty version of Twitter, namely Tweater. Users can post tweats with short recipes for delicious snacks. People can comment on these tweats. There are three tables: **users**, **tweats**, and **comments** that have relations among them. Which ones, you ask? You'll discover in a moment!

Let's start by importing the data on the users into your R session. You do this with the <a href="http://www.rdocumentation.org/packages/DBI/functions/dbReadTable" target="_blank" rel="noopener noreferrer">`dbReadTable()`</a> function. Simply pass it the connection object (`con`), followed by the name of the table you want to import. The resulting object is a standard R data frame.


**Instructions:**

* Add code that imports the `"users"` table from the tweater database and store the resulting data frame as `users`.
* Print the `users` data frame.

**Solution:**

```{r,eval=TRUE}
# Import the users table from tweater: users
users <- dbReadTable(con, "users")

# Print users
users
```



## Import all tables

Next to the `users`, we're also interested in the `tweats` and `comments` tables. However, separate `dbReadTable()` calls for each and every one of the tables in your database would mean a lot of code duplication. Remember about the `lapply()` function? You can use it again here! A connection is already coded for you, as well as a vector `table_names`, containing the names of all the tables in the database.


**Instructions:**

* Finish the `lapply()` function to import the `users`, `tweats` and `comments` tables in a single call. The result, a list of data frames, will be stored in the variable `tables`.
* Print `tables` to check if you got it right.

**Solution:**

```{r,eval=TRUE}
# Get table names
table_names <- dbListTables(con)

# Import all tables
tables <- lapply(table_names, dbReadTable, conn = con)

# Print out tables
tables
```



## How do the tables relate?

The connection to the MySQL database `con` has already been created for you. `tables`, a list containing the three tables as data frames that you've created in the previous exercise, is also available.

If you have a closer look at these tables, you'll see that the `tweats` table, for example, contains a column `user_id`. The ids in the column refer to the users that have posted the tweat. Similarly, the `comments` contain both a `user_id` and a `tweat_id` column. It specifies which user posted a comment on which tweat.

With this new knowledge, can you tell **who** posted the **tweat** on which somebody **commented** "awesome! thanks!" (comment 1012)?

**Possible Answers:**

:white_large_square: The user with id 1, so Kate.<br>
:white_large_square: There is not enough information to solve this.<br>
:white_large_square: The user with id 4, so Thomas.<br>
:white_check_mark: The user with user_id 5, so Oliver.<br>

# 2. Importing data from databases (Part 2)
## Query tweater (1)

In your life as a data scientist, you'll often be working with huge databases that contain tables with millions of rows. If you want to do some analyses on this data, it's possible that you only need a fraction of this data. In this case, it's a good idea to send SQL queries to your database, and only import the data you actually need into R.

<a href="http://www.rdocumentation.org/packages/DBI/functions/dbGetQuery" target="_blank" rel="noopener noreferrer">`dbGetQuery()`</a> is what you need. As usual, you first pass the connection object to it. The second argument is an SQL query in the form of a character string. This example selects the `age` variable from the `people` dataset where `gender` equals `"male"`:


```{r}
dbGetQuery(con, "SELECT age FROM people WHERE gender = 'male'")
```

A connection to the `tweater` database has already been coded for you.


**Instructions:**

* Use <a href="http://www.rdocumentation.org/packages/DBI/functions/dbGetQuery" target="_blank" rel="noopener noreferrer">`dbGetQuery()`</a> to create a data frame, `elisabeth`, that **selects** the `tweat_id` column **from** the `comments` table **where** elisabeth is the commenter, her `user_id` is 1
* Print out `elisabeth` so you can see if you queried the database correctly.

**Solution:**

```{r,eval=TRUE}
# Import tweat_id column of comments where user_id is 1: elisabeth
elisabeth <- dbGetQuery(con, "SELECT tweat_id FROM comments WHERE user_id = 1")

# Print elisabeth
elisabeth

```



## Query tweater (2)

Apart from checking equality, you can also check for *less than* and *greater than* relationships, with `<` and `>`, just like in R.

`con`, a connection to the `tweater` database, is again available.


**Instructions:**

* Create a data frame, `latest`, that **selects** the `post` column **from** the `tweats` table observations **where** the `date` is higher than `'2015-09-21'`.
* Print out `latest`.

**Solution:**

```{r,eval=TRUE}
# Import post column of tweats where date is higher than '2015-09-21': latest
latest <- dbGetQuery(con, "SELECT post
                           FROM tweats
                           WHERE date > \"2015-09-21\"")

# Print latest
latest
```



## Query tweater (3)

Suppose that you have a `people` table, with a bunch of information. This time, you want to find out the `age` and `country` of married males. Provided that there is a `married` column that's 1 when the person in question is married, the following query would work.


```{SQL}
SELECT age, country
  FROM people
    WHERE gender = "male" AND married = 1
```

Can you use a similar approach for a more specialized query on the `tweater` database?


**Instructions:**

* Create an R data frame, `specific`, that **selects** the `message` column **from** the `comments` table **where** the `tweat_id` is 77 **and** the `user_id` is greater than 4.
* Print `specific`.

**Solution:**

```{r,eval=TRUE}
# Create data frame specific
specific <- dbGetQuery(con, "SELECT message
                             FROM comments
                             WHERE tweat_id = 77 AND user_id > 4")

# Print specific
specific
```


## Query tweater (4)

There are also dedicated SQL functions that you can use in the `WHERE` clause of an SQL query. For example, `CHAR_LENGTH()` returns the number of characters in a string.


**Instructions:**

* Create a data frame, `short`, that **selects** the `id` and `name` columns **from** the `users` table **where** the number of characters in the `name` is strictly less than 5.
* Print `short`.

**Solution:**

```{r,eval=TRUE}
# Create data frame short
short <- dbGetQuery(con, "SELECT id, name
                          FROM users
                          WHERE CHAR_LENGTH(name) < 5")

# Print short
short
```



## Join the query madness!

Of course, SQL does not stop with the the three keywords `SELECT`, `FROM` and `WHERE`. Another very often used keyword is `JOIN`, and more specifically `INNER JOIN`. Take this call for example:


```{SQL}
SELECT name, post
  FROM users INNER JOIN tweats on users.id = user_id
    WHERE date > "2015-09-19"
```

Here, the `users` table is joined with the `tweats` table. This is possible because the `id` column in the `users` table corresponds to the `user_id` column in the `tweats` table. Also notice how `name`, from the `users` table, and `post` and `date`, from the `tweats` table, can be referenced to without problems.

Can you predict the outcome of the following query?


```{SQL}
SELECT post, message
  FROM tweats INNER JOIN comments on tweats.id = tweat_id
    WHERE tweat_id = 77
```

Feel free to experiment with the `con` object!

**Possible Answers:**

:white_large_square: Trying to get the results of this SQL query throws an error.<br>
:white_check_mark: A table with four observations, containing two columns: post and message.<br>
:white_large_square: A table with six observations, containing all columns in the tweats table.<br>
:white_large_square: A table with six observations, containing the columns post and message.<br>


## Send - Fetch - Clear

You've used <a href="http://www.rdocumentation.org/packages/DBI/functions/dbGetQuery" target="_blank" rel="noopener noreferrer">`dbGetQuery()`</a> multiple times now. This is a virtual function from the `DBI` package, but is actually implemented by the `RMySQL` package. Behind the scenes, the following steps are performed:


* Sending the specified query with <a href="http://www.rdocumentation.org/packages/DBI/functions/dbSendQuery" target="_blank" rel="noopener noreferrer">`dbSendQuery()`</a>;
* Fetching the result of executing the query on the database with <a href="http://www.rdocumentation.org/packages/DBI/functions/dbFetch" target="_blank" rel="noopener noreferrer">`dbFetch()`</a>;
* Clearing the result with <a href="http://www.rdocumentation.org/packages/DBI/functions/dbClearResult" target="_blank" rel="noopener noreferrer">`dbClearResult()`</a>.

Let's not use <a href="http://www.rdocumentation.org/packages/DBI/functions/dbGetQuery" target="_blank" rel="noopener noreferrer">`dbGetQuery()`</a> this time and implement the steps above. This is tedious to write, but it gives you the ability to fetch the query's result in chunks rather than all at once. You can do this by specifying the `n` argument inside <a href="http://www.rdocumentation.org/packages/DBI/functions/dbFetch" target="_blank" rel="noopener noreferrer">`dbFetch()`</a>.


**Instructions:**

* Inspect the <a href="http://www.rdocumentation.org/packages/DBI/functions/dbSendQuery" target="_blank" rel="noopener noreferrer">`dbSendQuery()`</a> call that has already been coded for you. It selects the comments for the users with an id above 4.
* Use <a href="http://www.rdocumentation.org/packages/DBI/functions/dbFetch" target="_blank" rel="noopener noreferrer">`dbFetch()`</a> twice. In the first call, import only two records of the query result by setting the `n` argument to `2`. In the second call, import all remaining queries (don't specify `n`). In both calls, simply print the resulting data frames.
* Clear `res` with <a href="http://www.rdocumentation.org/packages/DBI/functions/dbClearResult" target="_blank" rel="noopener noreferrer">`dbClearResult()`</a>.

**Solution:**

```{r,eval=TRUE}
# Send query to the database
res <- dbSendQuery(con, "SELECT * FROM comments WHERE user_id > 4")

# Use dbFetch() twice
dbFetch(res, n = 2)
dbFetch(res)

# Clear res
dbClearResult(res)
```



## Be polite and ...

Every time you connect to a database using <a href="http://www.rdocumentation.org/packages/DBI/functions/dbConnect" target="_blank" rel="noopener noreferrer">`dbConnect()`</a>, you're creating a new connection to the database you're referencing. `RMySQL` automatically specifies a maximum of open connections and closes some of the connections for you, but still: it's always polite to manually disconnect from the database afterwards. You do this with the <a href="http://www.rdocumentation.org/packages/DBI/functions/dbDisconnect" target="_blank" rel="noopener noreferrer">`dbDisconnect()`</a> function.

The code that connects you to the database is already available, can you finish the script?


**Instructions:**

* Using the technique you prefer, build a data frame `long_tweats`. It **selects** the `post` and `date` columns **from** the observations in `tweats` **where** the character length of the `post` variable exceeds `40`.
* Print `long_tweats`.
* Disconnect from the database by using <a href="http://www.rdocumentation.org/packages/DBI/functions/dbDisconnect" target="_blank" rel="noopener noreferrer">`dbDisconnect()`</a>.

**Solution:**

```{r,eval=TRUE}
# Create the data frame  long_tweats
long_tweats <- dbGetQuery(con, "SELECT post, date
                                FROM tweats
                                WHERE CHAR_LENGTH(post) > 40")

# Print long_tweats
print(long_tweats)

# Disconnect from the database
dbDisconnect(con)
```


# 3. Importing data from the web (Part 1)
## Import flat files from the web

In the video, you saw that the `utils` functions to import flat file data, such as <a href="http://www.rdocumentation.org/packages/utils/functions/read.table" target="_blank" rel="noopener noreferrer">`read.csv()`</a> and <a href="http://www.rdocumentation.org/packages/utils/functions/read.table" target="_blank" rel="noopener noreferrer">`read.delim()`</a>, are capable of automatically importing from URLs that point to flat files on the web.

You must be wondering whether Hadley Wickham's alternative package, `readr`, is equally potent. Well, figure it out in this exercise! The URLs for both a `.csv` file as well as a `.delim` file are already coded for you. It's up to you to actually import the data. If it works, that is…


**Instructions:**

* Load the `readr` package. It's already installed on DataCamp's servers.
* Use `url_csv` to read in the `.csv` file it is pointing to. Use the <a href="https://cran.r-project.org/web/packages/readr/readr.pdf" target="_blank" rel="noopener noreferrer">`read_csv()`</a> function. The `.csv` contains column names in the first row. Save the resulting data frame as `pools`.
* Similarly, use `url_delim` to read in the online `.txt` file. Use the <a href="https://cran.r-project.org/web/packages/readr/readr.pdf" target="_blank" rel="noopener noreferrer">`read_tsv()`</a> function and store the result as `potatoes`.
* Print `pools` and `potatoes`. Looks correct?

**Solution:**

```{r,eval=TRUE}
# Load the readr package
library(readr)

# Import the csv file: pools
url_csv <- "http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/swimming_pools.csv"

pools <- read_csv(url_csv)

# Import the txt file: potatoes
url_delim <- "http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/potatoes.txt"

potatoes <- read_tsv(url_delim)

# Print pools and potatoes
print(pools)
print(potatoes)
```



## Secure importing

In the previous exercises, you have been working with URLs that all start with `http://`. There is, however, a safer alternative to HTTP, namely HTTPS, which stands for HypterText Transfer Protocol Secure. Just remember this: HTTPS is relatively safe, HTTP is not.

Luckily for us, you can use the standard importing functions with `https://` connections since R version 3.2.2.


**Instructions:**

* Take a look at the URL in `url_csv`. It uses a secure connection, `https://`.
* Use <a href="http://www.rdocumentation.org/packages/utils/functions/read.table" target="_blank" rel="noopener noreferrer">`read.csv()`</a> to import the file at `url_csv`. The `.csv` file it is referring to contains column names in the first row. Call it `pools1`.
* Use <a href="https://cran.r-project.org/web/packages/readr/readr.pdf" target="_blank" rel="noopener noreferrer">`read_csv()`</a> to read in the same `.csv` file in `url_csv`. Call it `pools2`.
* Print out the structure of `pools1` and `pools2`. Looks like the importing went equally well as with a normal `http` connection!

**Solution:**

```{r,eval=TRUE}
# https URL to the swimming_pools csv file.
url_csv <- "https://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/swimming_pools.csv"

# Import the file using read.csv(): pools1
pools1 <- read.csv(url_csv)

# Import the file using read_csv(): pools2
pools2 <- read_csv(url_csv)

# Print the structure of pools1 and pools2
str(pools1)
str(pools2)
```



## Import Excel files from the web

When you learned about `gdata`, it was already mentioned that `gdata` can handle `.xls` files that are on the internet. `readxl` can't, at least not yet. The URL with which you'll be working is already available in the sample code. You will import it once using `gdata` and once with the `readxl` package via a workaround.


**Instructions:**

* Load the `readxl` and `gdata` packages. They are already installed on DataCamp's servers.
* Import the `.xls` file located at the URL `url_xls` using <a href="http://www.rdocumentation.org/packages/gdata/functions/read.xls" target="_blank" rel="noopener noreferrer">`read.xls()`</a> from `gdata`. Store the resulting data frame as `excel_gdata`.
* You can not use <a href="https://cran.r-project.org/web/packages/readxl/readxl.pdf" target="_blank" rel="noopener noreferrer">`read_excel()`</a> directly with a URL. Complete the following instructions to work around this problem:
* Use <a href="http://www.rdocumentation.org/packages/utils/functions/download.file" target="_blank" rel="noopener noreferrer">`download.file()`</a> to download the `.xls` file behind the URL and store it locally as `"local_latitude.xls"`.
* Call <a href="https://cran.r-project.org/web/packages/readxl/readxl.pdf" target="_blank" rel="noopener noreferrer">`read_excel()`</a> to import the local file, `"local_latitude.xls"`. Name the resulting data frame `excel_readxl`.

**Solution:**

```{r,eval=TRUE}
# Load the readxl and gdata package
library(readxl)
library(gdata)

# Specification of url: url_xls
url_xls <- "http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/latitude.xls"

# Import the .xls file with gdata: excel_gdata
excel_gdata <- read.xls(url_xls)

# Download file behind URL, name it local_latitude.xls
download.file(url_xls, "data//local_latitude.xls")

# Import the local .xls file with readxl: excel_readxl
excel_readxl <- read_excel("data//local_latitude.xls")
```



## Downloading any file, secure or not

In the previous exercise you've seen how you can read excel files on the web using the `read_excel` package by first downloading the file with the <a href="http://www.rdocumentation.org/packages/utils/functions/download.file" target="_blank" rel="noopener noreferrer">`download.file()`</a> function.

There's more: with <a href="http://www.rdocumentation.org/packages/utils/functions/download.file" target="_blank" rel="noopener noreferrer">`download.file()`</a> you can download any kind of file from the web, using HTTP and HTTPS: images, executable files, but also `.RData` files. An `RData` file is very efficient format to store R data.

You can load data from an `RData` file using the <a href="http://www.rdocumentation.org/packages/base/functions/load" target="_blank" rel="noopener noreferrer">`load()`</a> function, but this function does not accept a URL string as an argument. In this exercise, you'll first download the `RData` file securely, and then import the local data file.


**Instructions:**

* Take a look at the URL in `url_rdata`. It uses a secure connection, `https://`. This URL points to an `RData` file containing a data frame with some metrics on different kinds of wine.
* Download the file at `url_rdata` using <a href="http://www.rdocumentation.org/packages/utils/functions/download.file" target="_blank" rel="noopener noreferrer">`download.file()`</a>. Call the file `"wine_local.RData"` in your working directory.
* Load the file you created, `wine_local.RData`, using the <a href="http://www.rdocumentation.org/packages/base/functions/load" target="_blank" rel="noopener noreferrer">`load()`</a> function. It takes one argument, the path to the file, which is just the filename in our case. After running this command, the variable `wine` will automatically be available in your workspace.
* Print out the <a href="http://www.rdocumentation.org/packages/base/functions/summary" target="_blank" rel="noopener noreferrer">`summary()`</a> of the `wine` dataset.

**Solution:**

```{r,eval=TRUE}

# https URL to the wine RData file.
url_rdata <- "https://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/wine.RData"

# Download the wine file to your working directory
download.file(url_rdata, "data/wine_local.RData")

# Load the wine data into your workspace using load()
load("data/wine_local.RData")

# Print out the summary of the wine data
summary(wine)
```



## Reading a text file from the web

Wow, you've learned a lot of ways to import a data file from the web in the previous exercises. Let's see if you can remember what's possible and what's not.

Which way of importing  data is *NOT* possible?


**Possible Answers:**


:white_large_square: Importing a .csv file residing on the web using the URL with read.csv():
`read.csv("http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/swimming_pools.csv")`

:white_large_square: Downloading a remote excel and saving it to your working directory using download.file():
`download.file("http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/latitude.xlsx", "lat.xlsx")`

:white_large_square: Importing a .txt file residing on the web using the URL with read_tsv():
`read_tsv("http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/potatoes.txt")`

:white_check_mark: Using the load() function to load a remote RData file into the workspace with only the URL string:
`load("https://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/wine.RData"`


## HTTP? httr! (1)

Downloading a file from the Internet means sending a GET request and receiving the file you asked for. Internally, all the previously discussed functions use a GET request to download files.

`httr` provides a convenient function, <a href="http://www.rdocumentation.org/packages/httr/functions/GET" target="_blank" rel="noopener noreferrer">`GET()`</a> to execute this GET request. The result is a `response` object, that provides easy access to the status code, content-type and, of course, the actual content.

You can extract the content from the request using the <a href="http://www.rdocumentation.org/packages/httr/functions/content" target="_blank" rel="noopener noreferrer">`content()`</a> function. At the time of writing, there are three ways to retrieve this content: as a raw object, as a character vector, or an R object, such as a list. If you don't tell <a href="http://www.rdocumentation.org/packages/httr/functions/content" target="_blank" rel="noopener noreferrer">`content()`</a> how to retrieve the content through the `as` argument, it'll try its best to figure out which type is most appropriate based on the content-type.


**Instructions:**

* Load the `httr` package. It's already installed on DataCamp's servers.
* Use <a href="http://www.rdocumentation.org/packages/httr/functions/GET" target="_blank" rel="noopener noreferrer">`GET()`</a> to get the URL stored in `url`. Store the result of this `GET()` call as `resp`.
* Print the `resp` object. What information does it contain?
* Get the content of `resp` using <a href="http://www.rdocumentation.org/packages/httr/functions/content" target="_blank" rel="noopener noreferrer">`content()`</a> and set the `as` argument to `"raw"`. Assign the resulting vector to `raw_content`.
* Print the first values in `raw_content` with <a href="http://www.rdocumentation.org/packages/utils/functions/head" target="_blank" rel="noopener noreferrer">`head()`</a>.

**Solution:**

```{r,eval=TRUE}
# Load the httr package
library(httr)

# Get the url, save response to resp
url  <- "http://www.example.com/"
resp <- GET(url) 

# Print resp
resp

# Get the raw content of resp: raw_content
raw_content <- content(resp, as = "raw")

# Print the head of raw_content
head(raw_content)
```



## HTTP? httr! (2)

Web content does not limit itself to HTML pages and files stored on remote servers such as DataCamp's Amazon S3 instances. There are many other data formats out there. A very common one is JSON. This format  is very often used by so-called Web APIs, interfaces to web servers with which you as a client can communicate to get or store information in more complicated ways.

You'll learn about Web APIs and JSON in the video and exercises that follow, but some experimentation never hurts, does it?


**Instructions:**

* Use <a href="http://www.rdocumentation.org/packages/httr/functions/GET" target="_blank" rel="noopener noreferrer">`GET()`</a> to get the `url` that has already been specified in the sample code. Store the response as `resp`.
* Print `resp`. What is the content-type?
* Use <a href="http://www.rdocumentation.org/packages/httr/functions/content" target="_blank" rel="noopener noreferrer">`content()`</a> to get the content of `resp`. Set the `as` argument to `"text"`. Simply print out the result. What do you see?
* Use <a href="http://www.rdocumentation.org/packages/httr/functions/content" target="_blank" rel="noopener noreferrer">`content()`</a> to get the content of `resp`, but this time do not specify a second argument. R figures out automatically that you're dealing with a JSON, and converts the JSON to a named R list.

**Solution:**

```{r,eval=TRUE}
# Get the url
url <- "http://www.omdbapi.com/?apikey=72bc447a&t=Annie+Hall&y=&plot=short&r=json"

# Print resp
resp <- GET(url)
print(resp)

# Print content of resp as text
print(content(resp, as = "text"))

# Print content of resp
print(content(resp))
```


# 4. Importing data from the web (Part 2)
## From JSON to R

In the simplest setting, <a href="http://www.rdocumentation.org/packages/jsonlite/functions/fromJSON" target="_blank" rel="noopener noreferrer">`fromJSON()`</a> can convert character strings that represent JSON data into a nicely structured R list. Give it a try!

**Instructions:**

* Load the `jsonlite` package. It's already installed on DataCamp's servers.
* `wine_json` represents a JSON. Use `fromJSON()` to convert it to a list, named `wine`.
* Display the structure of `wine`


**Solution:**

```{r,eval=TRUE}
# Load the jsonlite package
library(jsonlite)

# wine_json is a JSON
wine_json <- '{"name":"Chateau Migraine", "year":1997, "alcohol_pct":12.4, "color":"red", "awarded":false}'

# Convert wine_json into a list: wine
wine <- fromJSON(wine_json)

# Print structure of wine
str(wine)
```



## Quandl API

As Filip showed in the video, <a href="http://www.rdocumentation.org/packages/jsonlite/functions/fromJSON" target="_blank" rel="noopener noreferrer">`fromJSON()`</a> also works if you pass a URL as a character string or the path to a local file that contains JSON data. Let's try this out on the Quandl API, where you can fetch all sorts of financial and economical data.


**Instructions:**

* `quandl_url` represents a URL. Use `fromJSON()` directly on this URL and store the result in `quandl_data`.
* Display the structure of `quandl_data`.

**Solution:**

```{r,eval=TRUE}
# Definition of quandl_url
quandl_url <- "https://www.quandl.com/api/v3/datasets/WIKI/FB/data.json?auth_token=i83asDsiWUUyfoypkgMz"

# Import Quandl data: quandl_data
quandl_data <- fromJSON(quandl_url)

# Print structure of quandl_data
str(quandl_data)
```



## OMDb API

In the video, you saw how easy it is to interact with an API once you know how to formulate requests. You also saw how to fetch all information on Rain Man from OMDb. Simply perform a <a href="http://www.rdocumentation.org/packages/httr/functions/GET" target="_blank" rel="noopener noreferrer">`GET()`</a> call, and next ask for the contents with the <a href="http://www.rdocumentation.org/packages/httr/functions/content" target="_blank" rel="noopener noreferrer">`content()`</a> function. This <a href="http://www.rdocumentation.org/packages/httr/functions/content" target="_blank" rel="noopener noreferrer">`content()`</a> function, which is part of the `httr` package, uses `jsonlite` behind the scenes to import the JSON data into R.

However, by now you also know that `jsonlite` can handle URLs itself. Simply passing the request URL to <a href="http://www.rdocumentation.org/packages/jsonlite/functions/fromJSON" target="_blank" rel="noopener noreferrer">`fromJSON()`</a> will get your data into R. In this exercise, you will be using this technique to compare the release year of two movies in the Open Movie Database.


**Instructions:**

* Two URLs are included in the sample code, as well as a `fromJSON()` call to build `sw4`. Add a similar call to build `sw3`.
* Print out the element named `Title` of both `sw4` and `sw3`. You can use the `$` operator. What movies are we dealing with here?
* Write an expression that evaluates to `TRUE` if `sw4` was released later than `sw3`. This information is stored in the `Year` element of the named lists.

**Solution:**

```{r,eval=TRUE}
# Definition of the URLs
url_sw4 <- "http://www.omdbapi.com/?apikey=72bc447a&i=tt0076759&r=json"
url_sw3 <- "http://www.omdbapi.com/?apikey=72bc447a&i=tt0121766&r=json"

# Import two URLs with fromJSON(): sw4 and sw3
sw4 <- fromJSON(url_sw4)
sw3 <- fromJSON(url_sw3)


# Print out the Title element of both lists
print(sw4$Title)
print(sw3$Title)

# Is the release year of sw4 later than sw3?
sw4$Year > sw3$Year
```



## JSON practice (1)

JSON is built on two structures: objects and arrays. To help you experiment with these, two JSON strings are included in the sample code. It's up to you to change them appropriately and then call `jsonlite`'s <a href="http://www.rdocumentation.org/packages/jsonlite/functions/fromJSON" target="_blank" rel="noopener noreferrer">`fromJSON()`</a> function on them each time.


**Instructions:**

* Change the assignment of `json1` such that the R vector after conversion contains the numbers 1 up to 6, in ascending order. Next, call <a href="http://www.rdocumentation.org/packages/jsonlite/functions/fromJSON" target="_blank" rel="noopener noreferrer">`fromJSON()`</a> on `json1`.
* Adapt the code for `json2` such that it's converted to a named list with two elements: `a`, containing the numbers 1, 2 and 3 and `b`, containing the numbers 4, 5 and 6. Next, call <a href="http://www.rdocumentation.org/packages/jsonlite/functions/fromJSON" target="_blank" rel="noopener noreferrer">`fromJSON()`</a> on `json2`.

**Solution:**

```{r,eval=TRUE}
# Challenge 1
json1 <- '[1, 2, 3, 4, 5, 6]'
fromJSON(json1)

# Challenge 2
json2 <- '{"a": [1, 2, 3], "b": [4, 5, 6]}'
fromJSON(json2)
```



## JSON practice (2)

We prepared two more JSON strings in the sample code. Can you change them and call `jsonlite`'s <a href="http://www.rdocumentation.org/packages/jsonlite/functions/fromJSON" target="_blank" rel="noopener noreferrer">`fromJSON()`</a> function on them, similar to the previous exercise?


**Instructions:**

* Remove characters from `json1` to build a 2 by 2 matrix containing only 1, 2, 3 and 4. Call <a href="http://www.rdocumentation.org/packages/jsonlite/functions/fromJSON" target="_blank" rel="noopener noreferrer">`fromJSON()`</a> on `json1`.
* Add characters to `json2` such that the data frame in which the json is converted contains an additional observation in the last row. For this observations, `a` equals 5 and `b` equals 6. Call <a href="http://www.rdocumentation.org/packages/jsonlite/functions/fromJSON" target="_blank" rel="noopener noreferrer">`fromJSON()`</a> one last time, on `json2`.

**Solution:**

```{r,eval=TRUE}
# Challenge 1
json1 <- '[[1, 2], [3, 4]]'
fromJSON(json1)

# Challenge 2
json2 <- '[{"a": 1, "b": 2}, {"a": 3, "b": 4}, {"a": 5, "b": 6}]'
fromJSON(json2)
```



## toJSON()

Apart from converting JSON to R with <a href="http://www.rdocumentation.org/packages/jsonlite/functions/fromJSON" target="_blank" rel="noopener noreferrer">`fromJSON()`</a>, you can also use <a href="http://www.rdocumentation.org/packages/jsonlite/functions/fromJSON" target="_blank" rel="noopener noreferrer">`toJSON()`</a> to convert R data to a JSON format. In its most basic use, you simply pass this function an R object to convert to a JSON. The result is an R object of the class `json`, which is basically a character string representing that JSON.

For this exercise, you will be working with a `.csv` file containing information on the amount of desalinated water that is produced around the world. As you'll see, it contains a lot of missing values. This data can be found on the URL that is specified in the sample code.


**Instructions:**

* Use a function of the `utils` package to import the `.csv` file directly from the URL specified in `url_csv`. Save the resulting data frame as `water`. Make sure that strings are *not* imported as factors.
* Convert the data frame `water` to a JSON. Call the resulting object `water_json`.
* Print out `water_json`.

**Solution:**

```{r,eval=TRUE}
# URL pointing to the .csv file
url_csv <- "http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/water.csv"

# Import the .csv file located at url_csv
water <- utils::read.csv(url_csv, stringsAsFactors = FALSE)

# Convert the data file according to the requirements
water_json <- toJSON(water)

# Print out water_json
print(water_json)
```



## Minify and prettify

JSONs can come in different formats. Take these two JSONs, that are in fact exactly the same: the first one is in a minified format, the second one is in a pretty format with indentation, whitespace and new lines:

```{json}
# Mini
{"a":1,"b":2,"c":{"x":5,"y":6}}

# Pretty
{
  "a": 1,
  "b": 2,
  "c": {
    "x": 5,
    "y": 6
  }
}
```

Unless you're a computer, you surely prefer the second version. However, the standard form that <a href="http://www.rdocumentation.org/packages/jsonlite/functions/fromJSON" target="_blank" rel="noopener noreferrer">`toJSON()`</a> returns, is the minified version, as it is more concise. You can adapt this behavior by setting the `pretty` argument inside <a href="http://www.rdocumentation.org/packages/jsonlite/functions/fromJSON" target="_blank" rel="noopener noreferrer">`toJSON()`</a> to `TRUE`. If you already have a JSON string, you can use <a href="http://www.rdocumentation.org/packages/jsonlite/functions/prettify" target="_blank" rel="noopener noreferrer">`prettify()`</a> or <a href="http://www.rdocumentation.org/packages/jsonlite/functions/prettify" target="_blank" rel="noopener noreferrer">`minify()`</a> to make the JSON pretty or as concise as possible.


**Instructions:**

* Convert the `mtcars` dataset, which is available in R by default, to a *pretty* `JSON`. Call the resulting JSON `pretty_json`.
* Print out `pretty_json`. Can you understand the output easily?
* Convert `pretty_json` to a minimal version using <a href="http://www.rdocumentation.org/packages/jsonlite/functions/prettify" target="_blank" rel="noopener noreferrer">`minify()`</a>. Store this version under a new variable, `mini_json`.
* Print out `mini_json`. Which version do you prefer, the pretty one or the minified one?

**Solution:**

```{r,eval=TRUE}
# Convert mtcars to a pretty JSON: pretty_json
pretty_json <- toJSON(mtcars, pretty = T)

# Print pretty_json
print(pretty_json)

# Minify pretty_json: mini_json
mini_json <- minify(pretty_json)

# Print mini_json
print(mini_json)
```


# 5. Importing data from statistical software packages
## Import SAS data with haven

`haven` is an extremely easy-to-use package to import data from three software packages: SAS, STATA and SPSS. Depending on the software, you use different functions:


* SAS: <a href="http://cran.r-project.org/web/packages/haven/haven.pdf" target="_blank" rel="noopener noreferrer">`read_sas()`</a>
* STATA: <a href="http://cran.r-project.org/web/packages/haven/haven.pdf" target="_blank" rel="noopener noreferrer">`read_dta()`</a> (or <a href="http://cran.r-project.org/web/packages/haven/haven.pdf" target="_blank" rel="noopener noreferrer">`read_stata()`</a>, which are identical)
* SPSS: <a href="http://cran.r-project.org/web/packages/haven/haven.pdf" target="_blank" rel="noopener noreferrer">`read_sav()`</a> or <a href="http://cran.r-project.org/web/packages/haven/haven.pdf" target="_blank" rel="noopener noreferrer">`read_por()`</a>, depending on the file type.

All these functions take one key argument: the path to your local file. In fact, you can even pass a URL; `haven` will then automatically download the file for you before importing it.

You'll be working with data on the age, gender, income, and purchase level (0 = low, 1 = high) of 36 individuals (Source: <a href="http://support.sas.com/learn/" target="_blank" rel="noopener noreferrer">SAS</a>). The information is stored in a SAS file, `sales.sas7bdat`, which is available in your current working directory. You can also download the data <a href="http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/sales.sas7bdat" target="_blank" rel="noopener noreferrer">here</a>.


**Instructions:**

* Load the `haven` package; it's already installed on DataCamp's servers.
* Import the data file `"sales.sas7bdat"`. Call the imported data frame `sales`.
* Display the structure of `sales` with <a href="http://www.rdocumentation.org/packages/utils/functions/str" target="_blank" rel="noopener noreferrer">`str()`</a>. Some columns represent categorical variables, so they should be factors.

**Solution:**

```{r,eval=TRUE}
# Load the haven package
library(haven)

# Import sales.sas7bdat: sales
sales <- read_sas("data/sales.sas7bdat")

# Display the structure of sales
str(sales)
```



## Import STATA data with haven

Next up are STATA data files; you can use <a href="http://cran.r-project.org/web/packages/haven/haven.pdf" target="_blank" rel="noopener noreferrer">`read_dta()`</a> for these.

When inspecting the result of the `read_dta()` call, you will notice that one column will be imported as a `labelled` vector, an R equivalent for the common data structure in other statistical environments. In order to effectively continue working on the data in R, it's best to change this data into a standard R class. To convert a variable of the class `labelled` to a factor, you'll need `haven`'s <a href="http://cran.r-project.org/web/packages/haven/haven.pdf" target="_blank" rel="noopener noreferrer">`as_factor()`</a> function.

In this exercise, you will work with data on yearly import and export numbers of sugar, both in USD and in weight. The data can be found at:
<a href="http://assets.datacamp.com/production/course_1478/datasets/trade.dta" target="_blank" rel="noopener noreferrer">http://assets.datacamp.com/production/course_1478/datasets/trade.dta</a>


**Instructions:**

* Import the data file directly from the URL using <a href="http://cran.r-project.org/web/packages/haven/haven.pdf" target="_blank" rel="noopener noreferrer">`read_dta()`</a>, and store it as `sugar`.
* Print out the structure of `sugar`. The `Date` column has class `labelled`.
* Convert the values in the `Date` column of `sugar` to dates, using `as.Date(as_factor(___))`.
* Print out the structure of `sugar` once more. Looks better now?

**Solution:**

```{r,eval=TRUE}
# Import the data from the URL: sugar
sugar <- read_dta("http://assets.datacamp.com/production/course_1478/datasets/trade.dta")

# Structure of sugar
str(sugar)

# Convert values in Date column to dates
sugar$Date <- as.Date(as_factor(sugar$Date))

# Structure of sugar again
str(sugar)
```



## What does the graphic tell?

A plot can be very useful to explore the relationship between two variables. If you pass the <a href="http://www.rdocumentation.org/packages/graphics/functions/plot" target="_blank" rel="noopener noreferrer">`plot()`</a> function two arguments, the first one will be plotted on the x-axis, the second one will be plotted on the y-axis.

The sugar trading data is again available at <a href="http://assets.datacamp.com/production/course_1478/datasets/trade.dta" target="_blank" rel="noopener noreferrer">http://assets.datacamp.com/production/course_1478/datasets/trade.dta</a>.

After you've imported the data frame, you should plot two of its variables, `Import` against `Weight_I`, and describe their relationship! `haven` is already loaded in your R session, so you can start importing straight away.


**Possible Answers:**

:white_large_square: The relation between the import figures in USD and the import figures in weight can be described with a quadratic function that has a local maximum.<br>
:white_check_mark: The import figures in USD and the import figures in weight are rather positively correlated.<br>
:white_large_square: The import figures in USD and the import figures in weight are negatively associated. Along the points there is a monotonic decreasing trend.<br>
:white_large_square: No relationship can be guessed between the weight and the value of the imported sugar.<br>


## Import SPSS data with haven

The `haven` package can also import data files from SPSS. Again, importing the data is pretty straightforward. Depending on the SPSS data file you're working with, you'll need either <a href="http://cran.r-project.org/web/packages/haven/haven.pdf" target="_blank" rel="noopener noreferrer">`read_sav()`</a> - for `.sav` files - or <a href="http://cran.r-project.org/web/packages/haven/haven.pdf" target="_blank" rel="noopener noreferrer">`read_por()`</a> - for `.por` files.

In this exercise, you will work with data on four of the Big Five personality traits for 434 persons (Source: <a href="http://staff.bath.ac.uk/pssiw/stats2/page16/page16.html" target="_blank" rel="noopener noreferrer">University of Bath</a>). The Big Five is a psychological concept including, originally, five dimensions of personality to classify human personality. The SPSS dataset is called <a href="http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/person.sav" target="_blank" rel="noopener noreferrer">`person.sav`</a> and is available in your working directory.


**Instructions:**

* Use <a href="http://cran.r-project.org/web/packages/haven/haven.pdf" target="_blank" rel="noopener noreferrer">`read_sav()`</a> to import the SPSS data in `"person.sav"`. Name the imported data frame `traits`.
* `traits` contains several missing values, or `NA`s. Run <a href="http://www.rdocumentation.org/packages/base/functions/summary" target="_blank" rel="noopener noreferrer">`summary()`</a> on it to find out how many `NA`s are contained in each variable.
* Print out a subset of those individuals that scored high on Extroversion *and* on Agreeableness, i.e. scoring higher than 40 on each of these two categories. You can use `subset()` for this.

**Solution:**

```{r,eval=TRUE}
# Import person.sav: traits
traits <- read_sav("data/person.sav")

# Summarize traits
summary(traits)

# Print out a subset
subset(traits, Extroversion > 40 & Agreeableness > 40)
```



## Factorize, round two

In the last exercise you learned how to import a data file using the command <a href="http://cran.r-project.org/web/packages/haven/haven.pdf" target="_blank" rel="noopener noreferrer">`read_sav()`</a>. With SPSS data files, it can also happen that some of the variables you import have the `labelled` class. This is done to keep all the labelling information that was originally present in the `.sav` and `.por` files. It's advised to coerce (or change) these variables to factors or other standard R classes.

The data for this exercise involves information on employees and their demographic and economic attributes (Source: <a href="http://cehd.gmu.edu/book/dimitrov/spss" target="_blank" rel="noopener noreferrer">QRiE</a>). The data can be found on the following URL:

<a href="http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/employee.sav" target="_blank" rel="noopener noreferrer">http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/employee.sav</a>


**Instructions:**

* Import the SPSS data straight from the URL and store the resulting data frame as `work`.
* Display the summary of the `GENDER` column of `work`. This information doesn't give you a lot of useful information, right?
* Convert the `GENDER` column in `work` to a factor, the class to denote categorical variables in R. Use <a href="http://cran.r-project.org/web/packages/haven/haven.pdf" target="_blank" rel="noopener noreferrer">`as_factor()`</a>.
* Once again display the summary of the `GENDER` column. This time, the printout makes much more sense.

**Solution:**

```{r,eval=TRUE}
# Import SPSS data from the URL: work
work <- read_sav("http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/employee.sav")

# Display summary of work$GENDER
summary(work$GENDER)

# Convert work$GENDER to a factor
work$GENDER <- as_factor(work$GENDER)

# Display summary of work$GENDER again
summary(work$GENDER)
```



## Import STATA data with foreign (1)

The `foreign` package offers a simple function to import and read *STATA* data: <a href="http://www.rdocumentation.org/packages/foreign/functions/read.dta" target="_blank" rel="noopener noreferrer">`read.dta()`</a>.

In this exercise, you will import data on the US presidential elections in the year 2000. The data in `florida.dta` contains the total numbers of votes for each of the four candidates as well as the total number of votes per election area in the state of Florida (Source: <a href="http://results.elections.myflorida.com/" target="_blank" rel="noopener noreferrer">Florida Department of State</a>). The file is available in your working directory, you can download it <a href="http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/florida.dta" target="_blank" rel="noopener noreferrer">here</a> if you want to experiment some more.


**Instructions:**

* Load the `foreign` package; it's already installed on DataCamp's servers.
* Import the data on the elections in Florida, `"florida.dta"`, and name the resulting data frame `florida`. Use <a href="http://www.rdocumentation.org/packages/foreign/functions/read.dta" target="_blank" rel="noopener noreferrer">`read.dta()`</a> without specifying extra arguments.
* Check out the last 6 observations of `florida` with `tail()`


**Solution:**

```{r,eval=TRUE}
# Load the foreign package
library("foreign")

# Import florida.dta and name the resulting data frame florida
florida <- read.dta("data/florida.dta")

# Check tail() of florida
tail(florida)
```



## Import STATA data with foreign (2)

Data can be very diverse, going from character vectors to categorical variables, dates and more. It's in these cases that the additional arguments of <a href="http://www.rdocumentation.org/packages/foreign/functions/read.dta" target="_blank" rel="noopener noreferrer">`read.dta()`</a> will come in handy.

The arguments you will use most often are `convert.dates`, `convert.factors`, `missing.type` and `convert.underscore`. Their meaning is pretty straightforward, as Filip explained in the video. It's all about correctly converting STATA data to standard R data structures. Type `?read.dta` to find out about about the default values.

The dataset for this exercise contains socio-economic measures and access to education for different individuals (Source: <a href="http://datatopics.worldbank.org/Gender/topic/education" target="_blank" rel="noopener noreferrer">World Bank</a>). This data is available as <a href="http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/edequality.dta" target="_blank" rel="noopener noreferrer">`edequality.dta`</a>, which is located in the `worldbank` folder in your working directory.


**Instructions:**

* Specify the path to the file using <a href="http://www.rdocumentation.org/packages/base/functions/file.path" target="_blank" rel="noopener noreferrer">`file.path()`</a>. Call it `path`. Remember the `"edequality.dta"` file is located in the `"worldbank"` folder.
* Use the `path` variable to import the data file in three different ways; each time show its structure with <a href="http://www.rdocumentation.org/packages/utils/functions/str" target="_blank" rel="noopener noreferrer">`str()`</a>:

* `edu_equal_1`: By passing only the file `path` to <a href="http://www.rdocumentation.org/packages/foreign/functions/read.dta" target="_blank" rel="noopener noreferrer">`read.dta()`</a>.
* `edu_equal_2`: By passing the file `path`, and setting `convert.factors` to `FALSE`.
* `edu_equal_3`: By passing the file `path`, and setting `convert.underscore` to `TRUE`.

**Solution:**

```{r,eval=TRUE}
# Specify the file path using file.path(): path
path <- file.path("data", "edequality.dta")

# Create and print structure of edu_equal_1
edu_equal_1 <- read.dta(path)
str(edu_equal_1)

# Create and print structure of edu_equal_2
edu_equal_2 <- read.dta(path, convert.factors = F)
str(edu_equal_2)


# Create and print structure of edu_equal_3
edu_equal_3 <- read.dta(path, convert.underscore = T)
str(edu_equal_3)
```



## Do you know your data?

The previous exercise dealt about socio-economic indicators and access to education of different individuals. The `edu_equal_1` dataset that you've built is already available in the workspace. Now that you have it in R, it's pretty easy to get some basic insights.

For example, you can ask yourself *how many observations (e.g. how many people) have an `age` higher than 40 and are `literate`?* When you call

```{r}
str(edu_equal_1)
```

You'll see that `age` is an integer, and `literate` is a factor, with the levels "yes" and "no". The following expression thus answers the question:


```{r}
nrow(subset(edu_equal_1, age > 40 & literate == "yes"))
```

Up to you to answer a similar question now:

*How many observations/individuals of Bulgarian ethnicity have an income above 1000?*

**Possible Answers:**

:white_large_square: 9457<br>
:white_large_square: 1000<br>
:white_check_mark: 8997<br>
:white_large_square: 10840<br>

## Import SPSS data with foreign (1)

All great things come in pairs. Where `foreign` provided <a href="http://www.rdocumentation.org/packages/foreign/functions/read.dta" target="_blank" rel="noopener noreferrer">`read.dta()`</a> to read Stata data, there's also <a href="http://www.rdocumentation.org/packages/foreign/functions/read.spss" target="_blank" rel="noopener noreferrer">`read.spss()`</a> to read SPSS data files. To get a data frame, make sure to set `to.data.frame = TRUE` inside <a href="http://www.rdocumentation.org/packages/foreign/functions/read.spss" target="_blank" rel="noopener noreferrer">`read.spss()`</a>.

In this exercise, you'll be working with socio-economic variables from different countries (Source: <a href="http://cw.routledge.com/textbooks/9780415372985/resources/datasets.asp" target="_blank" rel="noopener noreferrer">Quantative Data Analysis in Education</a>). The SPSS data is in a file called `international.sav`, which is in your working directory. You can also download it <a href="http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/international.sav" target="_blank" rel="noopener noreferrer">here</a> if you want to play around with it some more.


**Instructions:**

* Import the data file `"international.sav"` and have R convert it to a data frame. Store this data frame as `demo`.
* Create a boxplot of the `gdp` variable of `demo`.

**Solution:**

```{r,eval=TRUE}
# Import international.sav as a data frame: demo
demo <- read.spss("data/international.sav", to.data.frame = TRUE)

# Create boxplot of gdp variable of demo
boxplot(demo$gdp)
```



## Excursion: Correlation

If you're familiar with statistics, you'll have heard about Pearson's Correlation. It is a measurement to evaluate the linear dependency between two variables, say <mjx-container class="MathJax CtxtMenu_Attached_0" jax="CHTML" style="font-size: 116.7%; position: relative;" role="presentation" tabindex="0" ctxtmenu_counter="8"><mjx-math class="MJX-TEX" aria-hidden="true"><mjx-mi class="mjx-i"><mjx-c class="mjx-c1D44B TEX-I"></mjx-c></mjx-mi></mjx-math><mjx-assistive-mml role="presentation" unselectable="on" display="inline"><math xmlns="http://www.w3.org/1998/Math/MathML"><mi>X</mi></math></mjx-assistive-mml></mjx-container> and <mjx-container class="MathJax CtxtMenu_Attached_0" jax="CHTML" style="font-size: 116.7%; position: relative;" role="presentation" tabindex="0" ctxtmenu_counter="9"><mjx-math class="MJX-TEX" aria-hidden="true"><mjx-mi class="mjx-i"><mjx-c class="mjx-c1D44C TEX-I"></mjx-c></mjx-mi></mjx-math><mjx-assistive-mml role="presentation" unselectable="on" display="inline"><math xmlns="http://www.w3.org/1998/Math/MathML"><mi>Y</mi></math></mjx-assistive-mml></mjx-container>. It can range from -1 to 1; if it's close to 1 it means that there is a strong positive association between the variables. If <mjx-container class="MathJax CtxtMenu_Attached_0" jax="CHTML" style="font-size: 116.7%; position: relative;" role="presentation" tabindex="0" ctxtmenu_counter="10"><mjx-math class="MJX-TEX" aria-hidden="true"><mjx-mi class="mjx-i"><mjx-c class="mjx-c1D44B TEX-I"></mjx-c></mjx-mi></mjx-math><mjx-assistive-mml role="presentation" unselectable="on" display="inline"><math xmlns="http://www.w3.org/1998/Math/MathML"><mi>X</mi></math></mjx-assistive-mml></mjx-container> is high, also <mjx-container class="MathJax CtxtMenu_Attached_0" jax="CHTML" style="font-size: 116.7%; position: relative;" role="presentation" tabindex="0" ctxtmenu_counter="11"><mjx-math class="MJX-TEX" aria-hidden="true"><mjx-mi class="mjx-i"><mjx-c class="mjx-c1D44C TEX-I"></mjx-c></mjx-mi></mjx-math><mjx-assistive-mml role="presentation" unselectable="on" display="inline"><math xmlns="http://www.w3.org/1998/Math/MathML"><mi>Y</mi></math></mjx-assistive-mml></mjx-container> tends to be high. If it's close to -1, there is a strong negative association: If <mjx-container class="MathJax CtxtMenu_Attached_0" jax="CHTML" style="font-size: 116.7%; position: relative;" role="presentation" tabindex="0" ctxtmenu_counter="12"><mjx-math class="MJX-TEX" aria-hidden="true"><mjx-mi class="mjx-i"><mjx-c class="mjx-c1D44B TEX-I"></mjx-c></mjx-mi></mjx-math><mjx-assistive-mml role="presentation" unselectable="on" display="inline"><math xmlns="http://www.w3.org/1998/Math/MathML"><mi>X</mi></math></mjx-assistive-mml></mjx-container> is high, <mjx-container class="MathJax CtxtMenu_Attached_0" jax="CHTML" style="font-size: 116.7%; position: relative;" role="presentation" tabindex="0" ctxtmenu_counter="13"><mjx-math class="MJX-TEX" aria-hidden="true"><mjx-mi class="mjx-i"><mjx-c class="mjx-c1D44C TEX-I"></mjx-c></mjx-mi></mjx-math><mjx-assistive-mml role="presentation" unselectable="on" display="inline"><math xmlns="http://www.w3.org/1998/Math/MathML"><mi>Y</mi></math></mjx-assistive-mml></mjx-container> tends to be low. When the Pearson correlation between two variables is 0, these variables are possibly independent: there is no association between <mjx-container class="MathJax CtxtMenu_Attached_0" jax="CHTML" style="font-size: 116.7%; position: relative;" role="presentation" tabindex="0" ctxtmenu_counter="14"><mjx-math class="MJX-TEX" aria-hidden="true"><mjx-mi class="mjx-i"><mjx-c class="mjx-c1D44B TEX-I"></mjx-c></mjx-mi></mjx-math><mjx-assistive-mml role="presentation" unselectable="on" display="inline"><math xmlns="http://www.w3.org/1998/Math/MathML"><mi>X</mi></math></mjx-assistive-mml></mjx-container> and <mjx-container class="MathJax CtxtMenu_Attached_0" jax="CHTML" style="font-size: 116.7%; position: relative;" role="presentation" tabindex="0" ctxtmenu_counter="15"><mjx-math class="MJX-TEX" aria-hidden="true"><mjx-mi class="mjx-i"><mjx-c class="mjx-c1D44C TEX-I"></mjx-c></mjx-mi></mjx-math><mjx-assistive-mml role="presentation" unselectable="on" display="inline"><math xmlns="http://www.w3.org/1998/Math/MathML"><mi>Y</mi></math></mjx-assistive-mml></mjx-container>.

You can calculate the correlation between two vectors with the <a href="http://www.rdocumentation.org/packages/stats/functions/cor" target="_blank" rel="noopener noreferrer">`cor()`</a> function. Take this code for example, that computes the correlation between the columns `height` and `width` of a fictional data frame `size`:

```{r}
cor(size$height, size$width)
```

The data you've worked with in the previous exercise, <a href="http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/international.sav" target="_blank" rel="noopener noreferrer">international.sav</a>, is again available in your working directory. It's now up to import it and undertake the correct calculations to answer the following question:

*What is the correlation coefficient for the two numerical variables `gdp` and `f_illit` (female illiteracy rate)?*


**Possible Answers:**

:white_large_square: The correlation is very close to 0. Therefore, no association is existing between female illiteracy and GDP for the data set that is used.<br>
:white_check_mark: The correlation is around -0.45. There is a negative correlation, but it is rather weak.<br>
:white_large_square: The correlation is almost equal to +1. GDP and female illiteracy are almost perfectly, positive correlated.<br>
:white_large_square: The correlation is around +0.45. There is a positive correlation, but it is rather weak.<br>


## Import SPSS data with foreign (2)

In the previous exercise, you used the `to.data.frame` argument inside <a href="http://www.rdocumentation.org/packages/foreign/functions/read.spss" target="_blank" rel="noopener noreferrer">`read.spss()`</a>. There are many other ways in which to customize the way your SPSS data is imported.

In this exercise you will experiment with another argument, `use.value.labels`. It specifies whether variables with value labels should be converted into R factors with levels that are named accordingly. The argument is `TRUE` by default which means that so called labelled variables inside SPSS are converted to factors inside R.

You'll again be working with the <a href="http://s3.amazonaws.com/assets.datacamp.com/production/course_1478/datasets/international.sav" target="_blank" rel="noopener noreferrer">international.sav</a> data, which is available in your current working directory.


**Instructions:**

* Import the data file `"international.sav"` as a data frame, `demo_1`.
* Print the first few rows of `demo_1` using the <a href="http://www.rdocumentation.org/packages/utils/functions/head" target="_blank" rel="noopener noreferrer">`head()`</a> function.
* Import the data file `"international.sav"` as a data frame, `demo_2`, but this time in a way such that variables with value labels are *not* converted to R factors.
* Again, print the first few rows of `demo_2`. Can you tell the difference between the two data frames?

**Solution:**

```{r,eval=TRUE}
# foreign is already loaded

# Import international.sav as demo_1
demo_1 <- read.spss("data/international.sav", to.data.frame = T)

# Print out the head of demo_1
head(demo_1)

# Import international.sav as demo_2
demo_2 <- read.spss("data/international.sav", to.data.frame = T, use.value.labels = F)

# Print out the head of demo_2
head(demo_2)
```



